Wine Quality and Chemical Composition by Michael Jenkins
========================================================

> **Tip**: You will see quoted sections like this throughout the template to
help you construct your report. Make sure that you remove these notes before
you finish and submit your project!

> **Tip**: One of the requirements of this project is that your code follows
good formatting techniques, including limiting your lines to 80 characters or
less. If you're using RStudio, go into Preferences \> Code \> Display to set up
a margin line to help you keep track of this guideline!

A brief exploration of the relationship between wine quality and chemical 
composition.

# delete this line

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(knitr)
library(dplyr)
library(data.table)
library(GGally)
library(gridExtra)
library(reshape2)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wines.red <- read.csv("wineQualityReds.csv")
wines.white <- read.csv("wineQualityWhites.csv")
```

> **Tip**: Before you create any plots, it is a good idea to provide a short
introduction into the dataset that you are planning to explore. Replace this
quoted text with that general information!

The datasets explored in this analysis were found 
[here for red](https://bit.ly/2KTzkjX) and 
[here for white](https://bit.ly/1QqxDBL)

These datasets contain a sampling of wine chemical composition and a quality 
rating based on the opinion of at least three experts. The red wine sample 
contains 1,599 observations and the white wine sample contains 4898 
observations. Chemical composition is detailed in 11 variables.  

From these sample observations, we will explore the potential relationship 
between wine quality and chemical composition.  

A description of these datasets can be found [here](https://bit.ly/2vinuZz)  

## Citations

P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.  
Modeling wine preferences by data mining from physicochemical properties.  
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.  

[@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016  
[Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf  
[bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib  

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

```{r echo=FALSE, Univariate_Plots}
str(wines.red)
```

There's not much to go on with just raw numbers. We'll need to investigate the 
accompanying documentation to give context to these numbers. From the 
documentation we find the data set contains compositional measurements of 
various solutes found within a solution of wine. Also included is a measure of 
quality saved as an integer.  

This investigation is primarily concerned with predicting wine quality from 
wine composition. It only makes sense to start with an investigation of the 
quality score. How utilized is the 0-10 quality scale?

```{r echo=FALSE, Univariate_Plots}
summary(wines.red$quality)
```

With a minimum of 3 and a maximum of 8, I believe we can say the range is \
under utilized. There's also the potential for skew in the ordinal scale of \
quality. Given the median is near the mean, there's reason to be sceptical \
skew exists towards the extremes. What does give me pause is the 1st and 3rd \
quartiles are separated by 1 on an integer scale. This would suggest an \
inordinate number of values occur at the middling quality range. Histograms \
are cheap and will give a quick visualization to confirm.  

```{r echo=FALSE, Univariate_Plots}
ggplot(data=wines.red, aes(x=quality)) +
  geom_histogram(binwidth = 1)

```

> **Tip**: Make sure that you leave a blank line between the start / end of
each code block and the end / start of your Markdown text so that it is
formatted nicely in the knitted text. Note as well that text on consecutive
lines is treated as a single space. Make sure you have a blank line between
your paragraphs so that they too are formatted for easy readability.

It does appear that our dataset primarily contains wines of average quality. 
I find pie-charts are a good way to understand how homogenous a population is.

```{r echo=FALSE, Univariate_Plots}
red_quality_counts <- wines.red %>%
  group_by(quality) %>%
  summarise(no_obs = length(quality))

ggplot(data = red_quality_counts, 
       aes(x = "", y = no_obs, fill = quality)) + 
  geom_bar(width = 1, stat="identity") + 
  coord_polar("y", start=0)
```

Wines of average quality definitely account for a large portion of our sample. \
Without performing a calculation, I would say about 80% of the red wine sample \
are wines of average quality.  

It might be valuable to include the sample of white wines to increase the \
sample size.  

```{r echo=FALSE, Univariate_Plots}
str(wines.white)
```

The datasets share a source. It makes sense they would have compatible \
variables. However, white wine does contain about two and a half times the \
sample size of red wines. For the investigation, we should keep this in mind. \

The primary purpose of adding the white wine dataset is to hopefully increase \
the granularity of quality ratings. There is the possibility that including \
white wine will skew our data.


```{r echo=FALSE, Univariate_Plots}
summary(wines.white$quality)
```

Not much is different. The mean and 1st to 3rd quartiles are unchanged. On a \
more successful note, the quality range is better utilized. We now have a \
population of wines rated at 9.  

Perhaps the distribution of quality for white wines has greater utilization.  

```{r echo=FALSE, Univariate_Plots}
ggplot(data=wines.white, aes(x=quality)) +
  geom_histogram(binwidth = 1)

```

There is still a disproportionate number of average quality wines in the white 
sample.  

```{r echo=FALSE, Univariate_Plots}
white_quality_counts <- wines.white %>%
  group_by(quality) %>%
  summarise(no_obs = length(quality))

ggplot(data = white_quality_counts, 
       aes(x = "", y = no_obs, fill = quality)) + 
  geom_bar(width = 1, stat="identity") + 
  coord_polar("y", start=0)
```

That disproportion is more apparent in a pie chart. A lesser percentage of \
white wines are considered average quality. It seems approximately 75% of the \
sample is rated at 5 and 6 compared to the 80% of reds.  

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

There may not be enough of a sample size to predict if a wine is of \
exceptionally good or poor quality based on its composition alone. There does seem to be enough \
samples to determine the chemical make-up of an average quality wine.  

Perhaps it will be enough to predict if a wine is of average quality. If it \
doesn't, then it's either very bad or very good. With this thinking, analysis \
on average vs. not-average may provide the best insight this data-set can \
provide. We should be able to say what makes a wine average.  


### What is the structure of your dataset?

Some rudimentary research on wine research turned up \
[Wine Chemistry on Wikipedia](https://en.wikipedia.org/wiki/Wine_chemistry). \
From this we see there's a possibly of some missing measures. For example,\
phenolic compounds and proteins. The possibility of missing solute data is \
a cause of concern. However, The accompanying documentation makes the claim no \
compositional information is missing. We'll proceed with the information we do \
have.  

The datasets contain thirteen variables for each observation. Ten of those \
variables describe composition. Eight of the ten are measures of \
concentration. Alcohol content is provided as a percent. Density provides a \
measure of the entire solution. Some of the solute measures are the same. \
Some are of differing units. These will need to be converted to an equal units \
of measure. With equal units, we will be able to make compositional \
comparisons. 

The remaining variables are not compositional measurements. X is an index that \
is not linked to another dataset. It can be discarded. Quality is 
provided as an integer. There may be cause to convert it to a factor. \
Converting quality to a factor would reduce the scale. This may complicate \
and future additions to the dataset. Finally there's Ph. Ph is is a \
descriptor of acidity. It may be of interest to explore the relationship \
between total acidic composition and Ph.  

### What is/are the main feature(s) of interest in your dataset?

Of most interest are the measures of concentration. The weight per volume of \
each solute is assumed to have a corelative relationship to the wine's \
quality. Comparing the ratios of a wine's compostition correlated to the \
wine's quality is the most interesting feature to investigate.

### What other features in the dataset do you think will help support your \

Also of interest is the non-compositional value of Ph. There's a few questions \
we can try to answer here. What commonality between Ph values exists at varied \
concentrations of acidic solute? Is there a relationship between Ph and \
quality? 

With both red and white datasets, we can also investigate differences between \
the two types. How do their compositions differ? Do those compositional \
differences affect the qualitative scale between whites and reds? 

### Data Preparations

As previously noted, we will need to convert our compositional data to units \
of equal measure. What options are there and which option best fits our \
purpose? 

Our first option is weight per volume. The data is presented to us in this \
format. This could be a good measure to perform investigation. Grams per \
cubic meter may be a common measure to the drug markets. I'd like a unit of \
measure that is more widely understood. 

Parts per million (PPM) is a commonly reported unit of concentration. \
This would make for easily reported values. What effect would a scale in the \
millions have on visualations of the measurements?

The percentage of a solute within a solution should cover both of these \
concerns. A summation of solute percent can be used to determine the ratio of \
solvent in each solution of wine. Percents provide a more reasonable scale, \
and percents are commonly understood.

Having weighed the unit options and outlined some initial questions we can \
proceed with generating a working dataset. The following list should cover \
what has been considered so far.

1. Add a field to note the wine's color.
1. Merge the two datasets into one.
1. Add a logical field to note if the wine is ordinary. (i.e. quality is 5 or 6)
1. Convert density from grams per cubic centimeter to grams per cubic \
decimeter. (i.e. multiply by 1000)
1. Convert Sulfur dioxides from milligrams per cubic decimeter to grams per \
cubic decimeter. (i.e. divide by 1000)
1. Convert solute concentration values to percents. (i.e. divide their value \
by density and multiply by 100)
1. Add fields for total solute and solvent percents.

```{r echo=FALSE}

#quality <- factor(c(1:10))
wines.red <- read.csv("wineQualityReds.csv")
wines.white <- read.csv("wineQualityWhites.csv")

wines.red$color = "red"
wines.white$color = "white"

winesTmp <- rbind(wines.red,wines.white)

winesTmp$quality.class = ifelse((winesTmp$quality == 5 | 
                                   winesTmp$quality == 6), 'ordinary', 
                                'extreme')

winesTmp$quality.class <- factor(winesTmp$quality.class)
winesTmp$quality <- factor(winesTmp$quality)

# Drop unused index
winesTmp <- subset(winesTmp, select = -c(X))

# Convert values to grams per cubic decimeter
winesTmp$density <- winesTmp$density * 1000
winesTmp$free.sulfur.dioxide <- winesTmp$free.sulfur.dioxide / 1000
winesTmp$total.sulfur.dioxide <- winesTmp$total.sulfur.dioxide / 1000

# convert grams per cubic decimeter to percent solution
# load to a final dataframe
notConcentrate <- names(winesTmp) %in% 
  c('pH', 'color', 'alcohol', 'quality', 'quality.class', 'density')

wines <- winesTmp[,!notConcentrate] / winesTmp[,"density"] * 100
wines <- cbind(winesTmp[,notConcentrate],wines)

# validate total solution
solutes <- names(wines) %in% 
  c('alcohol', 'fixed.acidity', 'volatile.acidity', 'citric.acid', 
    'residual.sugar', 'chlorides', 'total.sulfur.dioxide', 'sulphates')

wines$total.solute <- rowSums(wines[,solutes])
wines$total.solvent <- 100 - wines$total.solute

rm('notConcentrate')
rm('winesTmp')
rm('wines.red')
rm('wines.white')

head(wines)

```

### Did you create any new variables from existing variables in the dataset?

The sum of solute percentages does not equal 100%. There is no information on \
the composition of the remaining solution. The information sheet claims there \
are no missing attribute values. Therefore we must assume the unaccounted \
solution is a tasteless solvent (e.g. water). Higher solvent content may be a \
factor in determining quality. We could hypothesize Watered down wine is of \
lower quality. Total percent solute and percent solvent were added for \
comparing these values.  

Another variable was added to retain information on which dataset the \
observation originated from. This will allow us to compare red wines to whites \
on their compositional differences and quality similarities.  

Including a boolean to say if a wine is average or exceptional was also added \
to reduce calculation times. In this case, exceptional is defined as not \
average. Exceptionally bad and exceptionally good wines have similar \
numbers of observations. Comparing what makes for an average vs. non-average \
wine and then comparing the extremes may give a more complete picture of wine \
quality.  

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Finally, the solute columns were changed from mass per volume to percents were \
renamed to better describe new unit of measure. Density was maintained in \
units of cubic decimeters. Density, by definition is 100% of the solution.  

# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

When cleaning up the dataset, a number of questions were raised on the \
relationships between variables. As a starting point we'll take a matrix \
synopsis between solutes and quality class

```{r echo=FALSE, Bivariate_Plots}

subjects <-   c('alcohol', 'fixed.acidity', 'volatile.acidity', 
    'citric.acid', 'residual.sugar', 'chlorides', 'total.sulfur.dioxide', 
    'sulphates', 'free.sulfur.dioxide', 'pH', 'quality.class', 'quality')

theme_set(theme_minimal(base_size=8))
ggpairs(wines[,subjects], progress = FALSE, lower=list(combo=wrap("facethist",
                                                                  binwidth=15)))

```

From this initial comparison, it seems alcohol content is the greatest \
predictor of wine quality. It also suggests the thought of average versus \
other will not give the results hoped for. Based on the distribution of \
alcohol content by quality, it makes more sense to bin quality as less than \
average and better than average values. The grouping would be split between 5 \
and 6.  

The strongest solute correlation is between free sulfur dioxide and total \
sulfur dioxide. This makes sense as one is a subset of the other. Sulfuric \
content related to quality does have an interesting phenomina. It appears that \
lower quality wines have larger range of sulfur dioxide content.

Acidic content also appears to have a slight impact on wine quality. The upper \
and lower bounds of acidic solutes lessens as quality improves. Total acidic \
content should be investigated further.

Total solvent and color are missing from the comparisons. They should be \
added in the next comparison set. Ordering of the variables should also \
adjusted for in the next comparison set.

```{R}
wines$quality.class = ifelse(as.numeric(wines$quality) <= 3, 'lesser', 'better')
wines$quality.class = factor(wines$quality.class, levels=c('lesser', 'better'))
wines$total.acid <- rowSums(wines[,c('fixed.acidity', 'volatile.acidity', 
                                     'citric.acid')])

subjects <- c('alcohol', 'total.solvent', 'pH', 'total.acid', 'fixed.acidity',
              'volatile.acidity','citric.acid', 'residual.sugar', 'chlorides', 
              'total.sulfur.dioxide', 'sulphates', 'free.sulfur.dioxide',
              'color', 'quality.class', 'quality')

theme_set(theme_minimal(base_size=8))
ggpairs(wines[,subjects], progress = FALSE, 
        lower=list(combo=wrap("facethist", binwidth=15)))
```

This new set has a higher number of correlated values. As alcohol content \
increases, total solvent decreases. This suggests that alcohol is a greater \
portion of the solution than other solutes. The opposite is true for \
total acid and fixed acidity. As one increases, so does the other. This \
suggests that fixed acidity is a higher proportion of total acidity than the \
other forms of acid in the solution.

Residual sugar appears to be consistent through all observations. This is \
surprising due to the relationship between sugar as a fuel for yeast to turn \
into alcohol.

Despite the low correlation between pH and total acidity, there does appear to \
be a loose relationship there. As total acidity increases, the pH drops \
(becomes more acidic). There's also appears to be a slight relationship \
between acidity and quality. Fixed acidity, citric acid and pH seems to be \
consistent through all quality grades. Volatile acidity has tends to have a \
higher content in lesser quality wines.

There's also greater variation in the chemical makeup between red and white \
wines. White wines tend to have more residual sugar, more sulfur dioxides, and \
a lower pH. The lower pH of white wines contrasts the lower acidic content of \
white wine. Remember, the lower the pH, the more acidic the compound.

From here, it would be good to compare red and white composition separately. \
Some variable limitation can also be done. Specifically residual sugar, \
chlorides, sulfur dioxides and sulfates can be discarded. Interest in those \
variables is zmainly in the compositional differences of red and white wines.

I did note earlier that the range of total sulfur dioxide is greater at lower \
qualities. However, the mean of sulfur dioxide content does remain constant. \
The difference in range could be attributed to white wine's tendancy to have a \
higher concentration of sulfur dioxides than red wines. Saving these \
values for a future analysis in the compositional differences between red \
and white wines would make sense.

The effect of acidic compounds and pH on quality also appears to have a \
dependency on wine variety. There's a noticable variation in acidic measures \
between red and white wines.

```{R}
subjects <- c('alcohol', 'total.solvent', 'pH', 'total.acid', 'fixed.acidity',
              'volatile.acidity','citric.acid', 'residual.sugar', 'chlorides', 
              'total.sulfur.dioxide', 'sulphates', 'free.sulfur.dioxide',
              'color', 'quality.class', 'quality')

theme_set(theme_minimal(base_size=8))
ggpairs(subset(wines[,subjects], color=="red"), progress = FALSE, 
        lower=list(combo=wrap("facethist", binwidth=15)))
```

```{R}
subjects <- c('alcohol', 'total.solvent', 'pH', 'total.acid', 'fixed.acidity',
              'volatile.acidity','citric.acid', 'residual.sugar', 'chlorides', 
              'total.sulfur.dioxide', 'sulphates', 'free.sulfur.dioxide',
              'color', 'quality.class', 'quality')

theme_set(theme_minimal(base_size=8))
ggpairs(subset(wines[,subjects], color=="white"), progress = FALSE, 
        lower=list(combo=wrap("facethist", binwidth=15)))
```

When splitting out the comparisons by color. It's easier to see how \
differences in composition can affect quality. Acidic and sulfuric compounds \
are a greater predictor of wine quality in reds than whites. Across both \
types, alcohol remains the greatest predictor of wine quality. Those compounds \
will be reserved for a future study on compositional differences between red \
and white wines.

Is there something that indirectly influences quality? A relationship between \
alcohol content and the remaining compounds? What are the ratios of other \
compounds compared to alcohol content?

Distribution of chlorides across alcohol seems to match the curve of alcohol \
content for all types. Less alcoholic wine tends to have more salt. Aside from \
some outliers, the percent salt content in red and whites appears to be \
similar.

Residual sugars content also appears to have a trend with alcohol content \
despite the low corelation. One factor that could account for this is starting \
sugar. The conversion rate of sugar to alcohol and interuptions in the \
fermentation process all contribute to the final alcohol and residual content.

The plots between alcohol, salt and sugar are worth exploring a bit closer. \
Before moving forward, it would help to take some steps back.

```{R alcohol counts} 
ggplot(aes(x = alcohol), data = wines) +
  geom_histogram(binwidth = .4, color = 'black', fill = 'orange')
```

Distribution of the alcohol content in our samples is right skewed. Are the \
samples of higher content skewing the quality contents? So far, we've \
been basing our thinking that alcohol content relates to wine quality by the \
upward trend of box plots at various levels of quality. We should take another \
view at the data to support this assumption.

```{R alcohol content for quality}
ggplot(data=wines, aes(x=quality, y=alcohol)) +
  geom_point()
```

There's quite a bit of overplotting in this graph

```{R alcohol content for quality}
ggplot(data=wines, aes(x=quality, y=alcohol)) +
  geom_jitter()
```

Wine samples graded as a 9 can be counted on one hand. There's still a bit of \
overplotting. The clusters of samples do seem to rise as quality improve.

```{R alcohol content for quality}
ggplot(data=wines, aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4) 
```

Reducing the alpha a bit and the clusters are more clear. It's also clear that \
we can't say that wine quality will be higher just because the alcohol \
content is higher. Higher quality wines do tend to have higher alcohol \
content. We are not saying that lower quality wines will always be of lower \
alcohol content.

```{R alcohol content for quality}
ggplot(data=wines, aes(x=quality.class, y=alcohol)) +
  geom_jitter(alpha=.4)
```

Taking the bins we've defined for lesser quality and better quality wines \
shows a clearer picture. Better quality wines are evenly dispersed across the \
range of alcohol content. It looks that the cut off is just above 10% alcohol \
content. Wines with more than 10% alcohol content are more likely to be of \
better quality.


Considering we might not want to completely give up the resolution on quality ,
it would be good to have another method of defining our lesser and better \
bins.

```{R alcohol content for quality}
wines %>%
  subset(as.numeric(quality) < 7) %>%
  group_by(class=cut(as.numeric(quality), breaks=seq(0, 7, by = 3))) %>%
  ggplot(data=., aes(x=class, y=alcohol)) +
  geom_jitter(alpha=.4)
```

We ended up needing to drop wines of quality 9, but this will be workable \
to maintain quality granularity while talking about wines of better vs lesser \
quality. I'm also regretting the decision to transform quality \
into a factor.

```{R alcohol content for quality}
wines %>% 
  subset(as.numeric(quality) > 3) %>%
  ggplot(aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4)
```

Taking a closer look at the breakdown of the better quality wines, We see that \
wines of quality 6 runs a full spectrum of alcohol content. Disregarding \
wines of quality 6, the alcohol content of better wines start to occur more \
frequently around 10-11%. Wines of even higher quality have even fewer \
occurances of alcohol content under ~10.5%

```{R alcohol content for quality}
wines %>% 
  subset(as.numeric(quality) >= 3 & as.numeric(quality) <= 5) %>%
  ggplot(aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4)
```

This shift for wines of greater vs lesser quality around 10% is more apparent \
if we limit our views to the quality scores surrounding where the limit should \
occur. Wines of quality 6 does appear to be a merge between wines of quality 5 \
and 7. The greatest concentration of quality 5 wines occurs under 10% alcohol \
content. The number of low quality wines is even more concentrated around \
~9.5% and lower.

```{R alcohol content for quality}
wines %>% 
  subset(as.numeric(quality) <= 3) %>%
  ggplot(aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4)
```

Breaking out the lesser quality wines by score shows there's an even \
dispersion under 11% alcohol content for wines of quality 3. For wines of \
quality 4, the dispersion is even upto about 11.5% quality.

If we were to eliminate wines with alcohol content between 9.5 and 11.5, what
would that do to our graphs? The greatest portion of our samples appear to \
fall in that range. Wines in that alcohol content range also appear to be \
make up the population of middling quality. 

```{R alcohol content for quality}
wines %>% 
  subset(alcohol <= 9.5 | alcohol >= 11.5) %>%
  ggplot(aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4) +
  stat_summary(fun.y = mean, fun.ymin = mean, fun.ymax = mean,
                 geom = "crossbar", width = 1, color = 'red')
```

The cut off between better and lesser quality wines is even easier to see when \
we remove the middle range of alcohol content. It does appear to occur at some \
granularity of a quality of 6.

As a quality 6 is ambiquous for determing lesser and better quality wines, \
we should look back to our lesser vs better quality graph and eliminate the \
ambiguous value.

```{R alcohol content for quality}
wines %>%
  subset(as.numeric(quality) < 7 & as.numeric(quality) !=4) %>%
  group_by(class=cut(as.numeric(quality), breaks=seq(0, 7, by = 3))) %>%
  ggplot(data=., aes(x=class, y=alcohol)) +
  geom_jitter(alpha=.4) +
  stat_summary(fun.y = mean, fun.ymin = mean, fun.ymax = mean,
                 geom = "crossbar", width = 1, color = 'red')
```

We should remember the quality scale is subjective based on the opinion of \
experts. It's possible those experts have a preference for higher alcohol \
content in their wines. On the observations made this far, I would suggest \
that preference is likely.

We should now take a look at actual values for the correlation between quality \
and alcohol.

```{R correlation}
cor.test(wines$alcohol, as.numeric(wines$quality))
```

Across all samples, there does seem to be a moderate correlation between \
quality and alcohol content. This is including a single value of wines quality \
that appears to contain both lesser and greater quality wines. An ambiguous \
measurement of quality. A value that fits the model but skews the data because \
of an apparent lack of resolution in the quality scale. What happens to \
correlation if we omit this ambiguous score?

```{R adjusted correlation}
wines.noMiddle <- subset(wines, as.numeric(quality) != 4)
cor.test(wines.noMiddle$alcohol, as.numeric(wines.noMiddle$quality))
rm(wines.noMiddle)
```

By omitting the ambiguous quality value on our scale, the correlation \
became strong. I would hypothesize, if we had a more granular and less \
subjective quality scale the correlation between would be even stronger.

What does this mean for the relationship between the objective alcohol content \
and the subjective quality scale? I would suggest that this supports the \
assumption that alcohol is fair objective measurement of quality. 

```{R alcohol summary}
summary(wines$alcohol)
```

Our earlier observations on the upper limit of alcohol content for lower \
quality wines and lower limit for higher quality wines are almost exactly the \
1st and 3rd quartiles of alcohol content ranges. This gives me more confidence \
in the assumption.

```{R adjusted correlation}
wines.noMiddle <- subset(wines, as.numeric(quality) != 4 & (alcohol >= 11.3 |
                                                              alcohol <= 9.5))
cor.test(wines.noMiddle$alcohol, as.numeric(wines.noMiddle$quality))
rm(wines.noMiddle)
```

Cutting out this additional measure of average wine produces an even stronger \
correlation between quality and alcohol content. 

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

To summarize, there is a strong correlation between wine quality and alcohol \
content for wines outside of average alcohol content and not of average \
quality. The number of qualifiers for this correlation is due to the \
granularity of the subjective scale used to convey quality. Without a more \
granular quality scale, we can only assume the trend maintains for wines \
of average quality and average alcohol content.

Given this assumption, we can take alcohol content as an objective measure of \
quality. With this as an objective measure, we can better investigate the \
ratios of other solutes as they relate to our quality surrogate.

Across both colors of wine, chlorides (salt) and residual sugar have an \
interesting relationship with alcohol. These compounds should be explored for \
their contribution to wine quality.


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

Using alcohol as a surrogate to quality, how does salt concentration compare \
at various levels of alcohol?

```{r echo=FALSE, Multivariate_Plots}
ggplot(data=wines, aes(x=alcohol, y=chlorides)) +
  geom_jitter(alpha=.4)
```

The majority of wines have less than a hundredth of a percent of salt. There's \
a jump in salt content around 9.5 percent alcohol. Most of them appear to be \
outliers. This could use some subsetting and scaling to get a better picture.

```{r echo=FALSE, Multivariate_Plots}
ggplot(data=subset(wines, chlorides < .03), aes(x=alcohol, y=chlorides)) +
  geom_jitter(alpha=.2) +
  scale_y_log10() +
  geom_smooth()
```

On a log10 scale, it's easier to see the fraction of a fractional difference \
between each level of alcohol content. Across the range of average alcohol \
content we see a consistent drop in salt content from ~0.006 to ~0.004 percent.  

```{R}
ggplot(data=subset(wines, chlorides < .03), aes(x=alcohol, y=chlorides)) +
  geom_jitter(alpha=.4) +
  scale_y_continuous(limits = c(.002, .006)) +
  geom_smooth()
```

Using this natural scale, the drop is even more dynamic.

What about sugar content?

```{R}
ggplot(data=wines, aes(x=alcohol, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_smooth()
```

Low alcohol content wine has a large amount of residual sugars. This seems \
logical as sugar is what yeast uses to create alcohol. There's interesting \
bumps in alcohol content near where the average and better than averge wines \
were determined. this should be compared to a scatterplot to look for any \
outliers that may be skewing the data.

```{R}
ggplot(data=wines, aes(x=alcohol, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_jitter(alpha=.5) +
  geom_smooth()
```

There does appear to be outliers above 2.5% residual sugar

```{R}
ggplot(data=subset(wines, residual.sugar <= 2.5), aes(x=alcohol, 
                                                  y=residual.sugar)) + 
  scale_y_continuous() +
  geom_jitter(alpha=.4) +
  geom_smooth()
```

The bump does seem to be valid. One reason this could be is the type of yeast \
used. Another reason could be interuption in the ferment process. Type of \
grape used is another possiblity. Grape type or wine color is one possiblity \
that we do have data to test.

```{R}
ggplot(data=subset(wines, residual.sugar <= 2.5), aes(x=alcohol, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_jitter(alpha=.4) +
  geom_smooth(aes(color=color))
```

Reds wines appear to have a consistent residual sugar content across all \
alcohol levels. It's white wines that have a sharp drop in sugar content \
across levels. Is there a similar phenomenon in salt content?

```{R}
ggplot(data=subset(wines, chlorides < .03), aes(x=alcohol, y=chlorides)) +
  geom_jitter(alpha=.4) +
  scale_y_log10(limits = c(.002, .02)) +
  geom_smooth(aes(color=color))
```

White wines have less salt than reds. White wines also have \
a more pronounced drop in salt content as alcohol content increases.

Returning to sugar content, sugar content in red wines is lower than \
white wines. It might be helpful to break these into their own scales.

```{R}
ggplot(data=subset(wines, residual.sugar <= 0.7 & color == "red"), aes(x=alcohol, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_jitter(alpha=.5) +
  geom_smooth(aes(color=color))
```

Unlike white wines, red wines appear to have a slight increase in sugar \
content as alcohol increases. It's so slight, starting around 0.225 and \
ending near 0.25. Given the low sample size and lack of measurement error, \
the best we can say is red wine has consistent sugar content across all \
values. There may be some miniscule differences, but there's not enough \
information to say sugar content is a contributing factor in wine quality.

To be sure, we can check this hypothesis.

```{R}
ggplot(data=subset(wines, residual.sugar <= 0.7 & color == "red"), 
       aes(x=quality, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_smooth(aes(color=color)) +
  geom_boxplot()
```

The majority of values reside under 0.4% sugar content. Adjusting both graphs \
might yield a better image. Before making this adjustment, it's worth noting \
the 1st quartile and mean is fairly consistent across quality range. There is \
some variation in the 3rd quartile, but no consistent pattern.  

```{R}
p1 <- ggplot(data=subset(wines, residual.sugar <= 0.4 & color == "red"),
             aes(x=alcohol, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_jitter(alpha=.5) +
  geom_smooth(aes(color=color))

p2 <- ggplot(data=subset(wines, residual.sugar <= 0.4 & color == "red"), 
       aes(x=quality, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_smooth(aes(color=color)) +
  geom_boxplot()


grid.arrange(p1,p2, ncol = 1)

```

Removing these outliers has produced a more consistent pattern in both \
graphs. The sugar content for the range of alcohol content is more flat. \
Sugar content at each quality score has a consistent mean with variation \
occuring at the 1st and 3rd quartiles. This also supports our earlier \
investigation showing that alcohol content is a good predictor of wine \
quality.  

One final graph I want to look at is alcohol content at ratios \
of salts and sugar.

```{R}
wine.tastes <- data.frame("sugar" = round(wines$residual.sugar, 2), 
                          "salt" = round(wines$chlorides, 4),
                          "alcohol" = wines$alcohol,
                          "type" = wines$color)
head(wine.tastes)

```

```{R}
wine.ratios <- melt(wine.tastes, id.vars = c("sugar", "salt", "type"))
wine.ratios
```

```{R}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100)) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))
```

What we see is a concentration of higher alcohol content at \
lower salt and lower sugar content. Some of that could be attributed to the \
intersection of red and white wines. 

Red wine tends to have a much lower sugar content and higher salt content than \
white wine. That does explain the lower leg of this graph.  

Another thing to consider is the range of alcohol content as wine quality. \
The spread of alcohol content across ratios of sugar to salt becomes more \
apparent by introducing more colors to our graph.

```{R}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradient2(low = "green", mid = "blue", high = "red", 
                       midpoint = 12) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))
```

It becomes even more clear that lower alcohol content wines tend to \
occur at extreme values of sugar and salt content.

```{R}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(
    c("orange","green","blue", "red"))(50)) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))
```

Introducing more colors doesn't provide any additional insight. It does help \
solidify what we've already seen. It also helps futher illustrate at which \
ratios of sugar to salt we're getting the highest alcohol content. There \
seems to be the highest concentration of red dots near 0.0025% salt and \
0.25% sugar.

It's worth noting that what has been refered to this far as a ratio between \
two compounds is a bit of a misnomer. It describes 100 grams of sugar for \
every gram of salt. We just see a higher conncentration of the highest alcohol \
content wines around 0.3% sugar and 0.003% salt. We're not seeing any examples \
at 2% sugar and 0.02% salt. There is no diagonal line that would describe a \
perfect ratio.

To be more accurate, we should say that the highest alcohol content wines \
occur between 0.003% and 0.004% salt content and 0.1 to 0.75% sugar content.

```{R}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("green","blue", "red"))(50)) +
  scale_y_continuous(limits = c(0,0.75)) + 
  scale_x_continuous(limits = c(0,0.006))
```

Zooming in on this range we can see that it isn't a true ratio. It's more of a \
happy spot where the balance between salt and sugar isn't too much in one \
direction or the other.

```{R}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradient2(low = "white", mid = "blue", high = "black", 
                   midpoint = 11.5) +
  scale_y_continuous(limits = c(0,0.75)) + 
  scale_x_continuous(limits = c(0,0.006))
```

Focusing on one color centered close to the lower middle range of the highest \
alcohol content and gradiated between white to black provides a good contrast \
on what's happening in the happy spot. We can better see where there's \
not enough sugar for the amount of salt in that bright blue. Just above that \
spot, we see a higher distribution of higher alcohol content wines in the \
darker colors. As sugar and salt gets a little too high, the alcohol content \
starts to sharply drop. It happens more rapidly for salt concentration than \
sugar concentration.

```{R}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradient2(low = "white", mid = "blue", high = "black", 
                   midpoint = 11.5) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))
```

Zooming back out with this color scale, it's easier to see the happy spot.  

This leaves the question of reds vs whites.  

```{R}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = subset(wine.ratios, wine.ratios$type == "red")) +
  geom_tile() +
  scale_fill_gradient2(low = "white", mid = "blue", high = "black", 
                   midpoint = 11.5) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))+
  geom_smooth(aes(color=type))
```

Red wine's contribution to the happy spot is clear and not unexpected.

```{R}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = subset(wine.ratios, wine.ratios$type == "white")) +
  geom_tile() +
  scale_fill_gradient2(low = "white", mid = "blue", high = "black", 
                   midpoint = 11.5) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))+
  geom_smooth(aes(color=type))
```

White wine's contribution to the happy spot is also quite clear. Although, \
white wine seems to be more forgiving on sugar content. Rather, white wines \
with high sugar content are fine. After an amount of salt, sugar becomes less \
acceptable.

```{R}
wine.quality <- data.frame("sugar" = round(wines$residual.sugar, 2), 
                          "salt" = round(wines$chlorides, 4),
                          "quality" = as.numeric(wines$quality),
                          "type" = wines$color)
head(wine.quality)
```

```{R}
wine.qratio <- melt(wine.quality, id.vars = c("sugar", "salt", "type"))
wine.qratio
```

```{R}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.qratio) +
  geom_tile() +
  scale_fill_gradient2(low = "white", mid = "blue", high = "black", 
                       midpoint = 5) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))+
  geom_smooth(aes(color=type))
```  
  
  The relationship between salt, sugar and quality doesn't change much when \
  reverting back from alcohol content to quality scores. Although, granularity \
  is lost. This strengthens our assumption that alcohol content is a good \
  measure of quality.
  
# Multivariate Analysis

Based on what we've seen, salt content has more impact on wine \
quality than sugar. Even with sugar rich white wines, salt content is the largest \
determining factor. In white wines, you can have more sugar in a higher \
quality wine as long as the salt content remains in a decreasing range.

Between red and white wines, there is a convergence where wines of the highest \
alcohol content occurs with a balance of sugar and salt. This sweet spot begs \
a question of wine quality and how it relates to alcohol. Is alcohol content \
just a by product of the process used to balance sugar with salt? 

Alcohol content as a granular measure of quality is something that shouldn't \
escape scrutiny. In this analysis, alcohol content was taken as a rough model \
for quality. It offered a level of granularity not present in the quality \
scale. One issue with this is it may hide the relationship of other solutes in \
the solution and how the combination of those solutes relate to quality.

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

```{R}
thm <- theme(plot.title = element_text(size = 16, face="bold"), 
            axis.title.x = element_text(size = 10, face="bold"),
            axis.title.y = element_text(size = 10, face="bold"),
            panel.background = element_rect(fill = "#c9d5f2", color = "darkgrey"),
            panel.grid.major = element_line(size = .5, linetype = "solid",
                                            color = "grey"),
            panel.grid.minor = element_line(size = .1, linetype = "dashed"),
            plot.background = element_rect(fill = "#e1e6f4"))
```

### Plot One
```{r echo=FALSE, Plot_One}
wines %>% 
  subset(alcohol <= 14.5) %>%
  ggplot(aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4, shape = 21, fill = I('#2679af')) +
  stat_summary(fun.y = mean, fun.ymin = function(z){quantile(z, 0.25)}, 
               fun.ymax = function(z){quantile(z, 0.75)},
                 geom = "crossbar", width = 1, fill = '#f70800', alpha = .3) +
  scale_y_continuous(breaks = seq(8, 14, 1)) +
  labs(title="Alcohol Content as Quality", y="Percent Alcohol Content",
       x = "Average Quality Score") + thm


```

### Description One

This data set has an issue of granularity within the quality scale. Without a \
a greater level of granularity, the deliniation between good an bad quality \
wines is lost. Alcohol content was identified as a surrogate to quality. It \
offers a greater level of granularity.

This granularity of quality found in alcohol content is illustrated here. The \
exceptionally good and exceptionally bad wines have a lack of observation \
density to produce consistency with this assumption. However, on the \
observations we do have, we see the alcohol content means of lesser quality \
wines are much lower than higher quality wines. Through the middling quality \
wines, where the majority of our observations lie, this trend is more \
pronounced.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(data=subset(wines, chlorides < .025), 
       aes(x=alcohol, y=(chlorides*100))) +
  geom_jitter(alpha=.3, shape = 21, fill = I('#605f5d')) +
  scale_y_continuous(limits=c(0.2, 1.2),breaks = seq(0.2, 1.2, 0.2)) +
  geom_smooth(data = subset(wines, color=="red"), aes(color = "Red")) +
  geom_smooth(data = subset(wines, color=="white"), aes(color = "White")) + 
  scale_colour_manual(name="Varietal", values=c("#540000", "#d2e8a9")) +
  labs(title="Salt and Alcohol Content", 
       y="Salt (ppm)", x = "Alcohol (ABV)") +
  thm
```

### Description Two

Using alcohol as a granular measurement of quality revealed salt \
content as a major factor for determing wine quality. Across both varities we \
see that as alcohol content increases, salt content decreases. It becomes \
apparent that better wines have less salt.

The amount of salt in the wine samples is miniscule. Where alcohol can be \
measured in full percents by volume, salt is best described in parts per \
million. It is interesting that slight variation in a small amount can have \
such a great effect on quality.

We also start to see a divergence between the two main varietals of wine. Red \
wines tend to have almost twice the amount of salt of white wines. This \
suggests that quality is reliant on a combination of compounds.


### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  #geom_tile() +
  geom_raster(aes(interpolate = TRUE))+
  scale_alpha_identity() +
  #scale_alpha_continuous() +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))+
  geom_smooth(data = subset(wine.ratios, type=="red"), aes(color = "Red")) +
  geom_smooth(data = subset(wine.ratios, type=="white"), aes(color = "White")) + 
  scale_colour_manual(name="Varietal", values=c("#540000", "#d2e8a9")) +
  thm



  scale_fill_gradient2(
    low = "#2679af", mid = "#3f0103", high = "black",
    #colours = colorRampPalette(c("#0176ba","#3f0103"))(3),
    name = "ABV",
    midpoint=12) +
```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!