# Chemical Composition as Predictor of Wine Quality

by Michael Jenkins  

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(knitr)
library(dplyr)
library(data.table)
library(GGally)
library(gridExtra)
library(reshape2)
library(RColorBrewer)
library(scales)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
wines.red <- read.csv("wineQualityReds.csv")
wines.white <- read.csv("wineQualityWhites.csv")
```

## Introduction

When shopping for a bottle of wine, it can be difficult to choose due to all \
of the available options. These datasets about wine could offer some insight \
to make those decisions easier. I have a preference for red wine. So that's \
where I'll begin.  

They contain a sampling of wine chemical composition and a quality rating. \
Quality is determined by an average of expert opinions. Compositional \
qualities are provided in a variety of measurements across 10 variables. In \
total, there are 6,497 observations with 1,599 red and 4,898 whites.  

This dataset should give a good exploration between the relationship of wine \
quality and chemical composition.  

The datasets explored in this analysis were found \
[here for red](https://bit.ly/2KTzkjX) and \
[here for white](https://bit.ly/1QqxDBL). The accompanying documentation of \
the datasets can be found [here](https://bit.ly/2vinuZz).  

### Citations

P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.  
Modeling wine preferences by data mining from physicochemical properties.  
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.  

[@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016  
[Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf  
[bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib  

## Initial Investigation

First a quick look at the data structure.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Initial_Data_Structure}
str(wines.red)
```

There's not much to go on with just raw numbers. We'll need to investigate the \
accompanying documentation to give context to these numbers. From the \
documentation we find the data set contains compositional measurements of \
various solutes found within a solution of wine. Also included is a measure of \
quality saved as an integer.  

This investigation is primarily concerned with predicting wine quality from \
wine composition. It only makes sense to start with an investigation of the \
quality score. How utilized is the 0-10 quality scale?  

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Scale_Validation}
summary(wines.red$quality)
```

With a minimum of three and a maximum of eight, it's fair to say the range is \
under utilized. With a median near the mean, there's no reason to suspect \
skew. What does give me pause is the 1st and 3rd quartiles are separated by 1. \
This would suggest an inordinate occurrences of middling quality range. \
Histograms are cheap and will give a quick visualization to confirm.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Red_Quality_Distribution}
ggplot(data=wines.red, aes(x=quality)) +
  geom_histogram(binwidth = 1)
```

It appears the dataset does primarily consist of wines of average quality. \
A pie-chart would provide an understanding of how homogenous the population \
quality is.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
red_quality_counts <- wines.red %>%
  group_by(quality) %>%
  summarise(no_obs = length(quality))

ggplot(data = red_quality_counts, 
       aes(x = "", y = no_obs, fill = quality)) + 
  geom_bar(width = 1, stat="identity") + 
  coord_polar("y", start=0)
```

Wines of average quality definitely account for a large portion of our sample. \
Without performing a calculation, I estimate about 80% of the red wine sample \
are of average quality.  

It might be valuable to include the sample of white wines to increase the \
sample size.  

```{r echo=FALSE, message=FALSE, warning=FALSE, White_Wine_Structure}
str(wines.white)
```

The datasets share a source. It makes sense they would have compatible \
variables. However, white wine does contain about two and a half times the \
sample size of red wines.  

Remember, the primary purpose of adding the white wine dataset to the \
investigation is to increase the granularity of quality ratings. There is the \
possibility that including white wine will skew our data.  

```{r echo=FALSE, message=FALSE, warning=FALSE, White_Wine_Quality_Scale}
summary(wines.white$quality)
```

Not much is different. The mean and 1st to 3rd quartiles are unchanged. On a \
successful note, the quality range is better utilized. We now have a \
population of wines rated as nine.  

Perhaps the distribution of quality for white wines has greater utilization.  

```{r echo=FALSE, message=FALSE, warning=FALSE, White_Wine_Quality_Distribution}
ggplot(data=wines.white, aes(x=quality)) +
  geom_histogram(binwidth = 1)
```

That's diappointing. There is still a disproportionate number of average \
quality wines in the white sample.  

```{r echo=FALSE, message=FALSE, warning=FALSE, White_Wine_Quality_Proportions}
white_quality_counts <- wines.white %>%
  group_by(quality) %>%
  summarise(no_obs = length(quality))

ggplot(data = white_quality_counts, 
       aes(x = "", y = no_obs, fill = quality)) + 
  geom_bar(width = 1, stat="identity") + 
  coord_polar("y", start=0)
```

That disproportion is more apparent in a pie chart. Still, A lesser percentage \
of white wines are considered average quality. It seems approximately 75% of \
the sample is rated at five and six compared to the 80% of reds. So there \
was some success in our attempt at gaining granularity.  

### What Next

There may not be enough of a sample size to predict if a wine is of \
exceptionally good or poor quality based on its composition alone. There does \
seem to be enough samples to determine the chemical make-up of an average \
quality wine.  

Perhaps it will be enough to predict if a wine is of average quality. If it \
doesn't, then it's either very bad or very good. With this thinking, analysis \
on average vs. not-average may provide the best insight this data-set can \
provide. We might be able to say what makes a wine average.  

#### Delving into the Dataset

Some rudimentary research on wine research turned up \
[Wine Chemistry on Wikipedia](https://en.wikipedia.org/wiki/Wine_chemistry). \
From this we see there's a possibly of some missing measures. For example, \
phenolic compounds and proteins. The possibility of missing solute data is \
a cause of concern. However, The accompanying documentation makes the claim no \
compositional information is missing. We'll proceed with the information we do \
have.  

The datasets contain thirteen variables for each observation. Ten of those \
variables describe composition. Eight of the ten are measures of \
concentration. Alcohol content is provided as a percent. Density provides a \
measure of the entire solution. Some of the solute measures are the same. \
Some are of different units. These should be converted to equal units of \
measure. With equal units, we can make better compositional comparisons.  

The remaining variables are not compositional measurements. X is an index that \
is not linked to another dataset. It can be discarded. Quality is 
provided as an integer. There may be cause to convert it to a factor. \
Converting quality to a factor would reduce the scale. This may complicate \
any future additions to the dataset. Finally there's Ph. Ph is is a \
descriptor of acidity. It may be of interest to explore the relationship \
between total acidic composition and Ph.  

#### Points of Interest

Of most interest are the measures of concentration. The mass per volume of \
each solute is assumed to have a corelative relationship to the wine's \
quality. Comparing the ratios of a wine's compostition correlated to the \
wine's quality is the most interesting feature to investigate.  

Another interest is the non-compositional value of Ph. There's a few questions \
we can try to answer here. What commonality between Ph values exists at varied \
concentrations of acidic solute? Is there a relationship between Ph and \
quality?  

With both red and white datasets, we can also investigate differences between \
the two types. How do their compositions differ? Do those compositional \
differences affect the qualitative scale between whites and reds?  

## Data Preparation

As previously noted, we will need to convert our compositional data to units \
of equal measure. First we should consider what options are available and \
which best fits our purpose.  

The first option is mass per volume. The data is presented to us in this \
format. This could be a good measure to perform investigation. Grams per \
cubic meter may be a common measure to the drug markets. I'd like a unit of \
measure that is more widely understood.  

Parts per million (ppm) is a commonly reported unit of concentration. This \
would make for easily reported values. What effect would a scale in the \
millions have on visualizations?

A solute's percentage of a solution should cover these concerns. A summation \
of solute percent can be used to determine the ratio of solvent in each \
solution of wine. Percents should provide a more manageable scale, and are \
commonly understood.

If any of these assumptions about percents proves false, it will be easy to \
convert back to another better suited measure. 1% is equal to 10,000 ppm and 1 \
ppm is equal to 1 mg/L. We can also easily convert to grams per serving. \
There's 33.8 ounces in a liter and 5 ounces in a serving.

Having considered the options, we can generate a working dataset. The \
following list should cover what has been decided on.

1. Add a field to note the wine's color.
1. Merge the two datasets into one.
1. Add a logical field to note if the wine is ordinary. (i.e. quality is 5 or 6)
1. Convert density from grams per cubic centimeter to grams per cubic \
decimeter. (i.e. multiply by 1000)
1. Convert Sulfur dioxides from milligrams per cubic decimeter to grams per \
cubic decimeter. (i.e. divide by 1000)
1. Convert solute concentration values to percents. (i.e. divide their value \
by density and multiply by 100)
1. Add fields for total solute and solvent percents.

```{r echo=FALSE, message=FALSE, warning=FALSE, Data_Conversion}
# Add a color field
wines.red$color = "red"
wines.white$color = "white"

# Merge the datasets into a temporary table
winesTmp <- rbind(wines.red,wines.white)

# Add a field to delineate average wines from others.
winesTmp$quality.class = ifelse((winesTmp$quality == 5 | 
                                   winesTmp$quality == 6), 'ordinary', 
                                'extreme')

# Factor the quality values
winesTmp$quality.class <- factor(winesTmp$quality.class)
winesTmp$quality <- factor(winesTmp$quality)

# Drop the unused index
winesTmp <- subset(winesTmp, select = -c(X))

# Convert values to grams per cubic decimeter
winesTmp$density <- winesTmp$density * 1000
winesTmp$free.sulfur.dioxide <- winesTmp$free.sulfur.dioxide / 1000
winesTmp$total.sulfur.dioxide <- winesTmp$total.sulfur.dioxide / 1000

# Group columns that will be excluded in percent conversion
notConcentrate <- names(winesTmp) %in% 
  c('pH', 'color', 'alcohol', 'quality', 'quality.class', 'density')

# Convert grams per cubic decimeter to percent solution 
# Store results in final table
wines <- winesTmp[,!notConcentrate] / winesTmp[,"density"] * 100
wines <- cbind(winesTmp[,notConcentrate],wines)

# Prepare columns for solute to solvent calculations
solutes <- names(winesTmp) %in% 
  c('alcohol', 'fixed.acidity', 'volatile.acidity', 'citric.acid', 
    'residual.sugar', 'chlorides', 'total.sulfur.dioxide', 'sulphates')

# Calculate and add solute and solvent percents to table
wines$total.solute <- rowSums(wines[,as.numeric(solutes)])
wines$total.solvent <- 100 - wines$total.solute

# clean up after ourselves
rm('notConcentrate')
rm('winesTmp')
rm('wines.red')
rm('wines.white')
```

### Data Preparation Aftermath

Having cleaned up the data, I'd like to take note of the variables added. It \
is good to keep track of such things. Just in case we forget why they were \
added.  

The sum of the percent solutes does not account for the total solution. With \
no information on the composition of the remaining solution, and the claim \
there are no missing attributes, we must assume the unaccounted solution is a \
tasteless solvent (e.g. water). We're also making the assumption that alcohol \
is not a solvent for this case. Higher solvent content may be a factor in \
determining quality. We could hypothesize Watered down wine is of lower \
quality. These values will allow us to test these hypotheses.  

Another variable we added retains information on which dataset the observation \
originated from. This will allow us to compare the differences between red and \
white wines. As someone with a preference for reds, I think this part is the \
most important.  

The exceptional vs ordinary variable is intended for exploring what is common \
in wines of average quality. In this case, exceptional is defined as not \
average. Exceptionally bad and exceptionally good wines have similar numbers \
of observations. Comparing what makes for average vs. non-average wine and \
then comparing both ends of the extremes may give a more complete picture of \
wine quality.  

Finally, the solute columns were changed from mass per volume to percents. \
Density, by definition is 100% of the solution this is shown in our \
calculation. We should also not forget the unit conversions we have. While we \
are working with percents, it helps to be able to report whichever value is \
most conducive.

1% is equal to:  

* 10,000 ppm
* 10 g/L
* 1.48 g per serving

## Deeper Investigation

When cleaning up the dataset, a number of questions were raised on the \
relationships between variables. A matrix synopsis between the solutes should \
give a good direction on where to go.  

### Universal Comparisons

```{r echo=FALSE,  message=FALSE, warning=FALSE,The_Matrix}
# Define items for the matrix
subjects <- c('alcohol', 'fixed.acidity', 'volatile.acidity', 
    'citric.acid', 'residual.sugar', 'chlorides', 'total.sulfur.dioxide', 
    'sulphates', 'free.sulfur.dioxide', 'pH', 'quality.class', 'quality')

# Create the matrix
theme_set(theme_minimal(base_size=8))
ggpairs(wines[,subjects], progress = FALSE, 
        lower=list(combo=wrap("facethist", binwidth=15)))
```

From this initial comparison, it seems alcohol content is the best predictor \
of wine quality. It also suggests the thought of average versus other may not \
provide the results hoped for. Based on the distribution of alcohol content by \
quality, there's more sense in breaking quality as less than average and \
better than average values. This would regroup the observations with a split \
between qualities 5 and 6.  

The strongest solute correlation is between free sulfur dioxide and total \
sulfur dioxide. This makes sense as one is a subset of the other. Sulfuric \
content related to quality does have an interesting phenomena. It appears that \
lower quality wines have larger range of sulfur dioxide content.

Acidic content also appears to have a slight impact on wine quality. The upper \
and lower bounds of acidic solutes lessens as quality improves. Total acidic \
content could be investigated further.

Total solvent and color are missing from the comparisons. They should be added \
in the next comparison set. Variable ordering should also be adjusted in the \
next comparison set.

### Refined Comparisons

```{r echo=FALSE, message=FALSE, warning=FALSE, The_Matrix_Reloaded}
# Add new values
wines$quality.class = ifelse(as.numeric(wines$quality) <= 3, 'lesser', 'better')
wines$quality.class = factor(wines$quality.class, levels=c('lesser', 'better'))
wines$total.acid <- rowSums(wines[,c('fixed.acidity', 'volatile.acidity', 
                                     'citric.acid')])

# Redefine values for the matrix
subjects <- c('alcohol', 'total.solvent', 'pH', 'total.acid', 'fixed.acidity',
              'volatile.acidity','citric.acid', 'residual.sugar', 'chlorides', 
              'total.sulfur.dioxide', 'sulphates', 'free.sulfur.dioxide',
              'color', 'quality.class', 'quality')

# Reload the matrix
theme_set(theme_minimal(base_size=8))
ggpairs(wines[,subjects], progress = FALSE, 
        lower=list(combo=wrap("facethist", binwidth=15)))
```

This new set has a higher number of correlated values. As alcohol content \
increases, total solvent decreases. This suggests that alcohol is a greater \
portion of the solution than the other solutes. The opposite is true for \
total acid and fixed acidity. As one increases, so does the other. This \
suggests that fixed acidity has a higher proportion of total acidity than the \
other acidic forms.  

Residual sugar appears to be consistent through all observations. This is \
surprising due to the relationship between sugar as a fuel for yeast to turn \
into alcohol.  

Despite the low correlation between pH and total acidity, there does appear to \
be a loose relationship. As total acidity increases, the pH drops (becomes \
more acidic). There's also appears to be a slight relationship between acidity \
and quality. Fixed acidity, citric acid and pH seems to be consistent through \
all grade qualities. Volatile acidity seems to have a higher content in lesser \
quality wines.  

We start to see hints at the variation in chemical makeup between red and \
white wines. White wines tend to have more residual sugar, more sulfur \
dioxides, and lower pH. The lower pH of white wines contrasts with the lower \
acidic content of white wine. Remember, lower pH means more acidic. The \
expectation is higher acidic content would have lower pH. One possible \
explanation is fixed vs. other forms of acid.  

I did note earlier that the range of total sulfur dioxide is greater at lower \
qualities. However, the mean of sulfur dioxide content does remain constant. \
The difference in range could be attributed to white wine's tendency to have a \
higher concentration of sulfur dioxides than red wines. There's some logic to \
save these values for a future analysis. One with greater focus on the \
compositional differences of red and white wines.  

The effect of acidic compounds and pH on quality also appears to be dependent \
on wine variety. With all of these differences between red and white, It would \
be good to compare red and white composition separately.  

### Comparisons of Varietals

#### Red

```{r echo=FALSE, message=FALSE, warning=FALSE, The_Matrix_Red_Band}
theme_set(theme_minimal(base_size=8))
ggpairs(subset(wines[,subjects], color=="red"), progress = FALSE, 
        lower=list(combo=wrap("facethist", binwidth=15)))
```

#### White

```{r echo=FALSE, message=FALSE, warning=FALSE, The_Matrix_White_Band}
theme_set(theme_minimal(base_size=8))
ggpairs(subset(wines[,subjects], color=="white"), progress = FALSE, 
        lower=list(combo=wrap("facethist", binwidth=15)))
```

When splitting out the comparisons by color. It's easier to see how \
differences in composition affects quality. Acidic and sulfuric compounds are \
a greater predictor of wine quality in reds than whites. Across both types, \
alcohol remains the greatest predictor of wine quality. Acidic and sulfuric \
compounds will be reserved for future analysis.  

Is there something that indirectly influences quality? A relationship between \
alcohol content and the remaining compounds? What are the ratios of other \
compounds compared to alcohol content?  

Distribution of chlorides (salt) across alcohol seems to match the curve of \
alcohol content for all types. Less alcoholic wine tends to have more salt. \
Aside from some outliers, the percent salt content in red and whites appears \
to be similar.

Despite the low correlation, residual sugar and alcohol content also appears \
to have a universal trend across the varietals. We can only speculate on the \
unknown factors that would produce such a result. One factor could be the \
starting sugar. Another might be the yeast used to convert sugar to alcohol. \
At what alcohol content does that yeast die? Was the fermentation process \
interrupted at a certain point? These are questions we don't have the answers \
too. We shouldn't ignore this apparent relationship due to unknown influences.  

From all of this, the plots between alcohol, salt and sugar should be the \
center of our exploration. There appears to be some type of relationship \
between alcohol and quality. There's also appears to be a relationship between \
ratios of salt and sugar to alcohol. Before moving forward, it's only prudent \
to take some steps back.  

### Alcohol and Quality

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_Content_Distribution} 
ggplot(aes(x = alcohol), data = wines) +
  geom_histogram(binwidth = .4, color = 'black', fill = 'orange')
```

Distribution of the alcohol content in our samples is right skewed. Are the \
samples of higher content skewing the quality contents? So far, we've \
been directing our thinking on alcohol content and wine quality by the \
tendency of higher average alcohol content at higher levels of quality. We \
should take a closer view at the data to support this assumption.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_by_Quality}
ggplot(data=wines, aes(x=quality, y=alcohol)) +
  geom_point()
```

There's quite a bit of overplotting in that graph. Still, there is some \
pattern in the extreme qualities.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_by_Quality_Visibility}
ggplot(data=wines, aes(x=quality, y=alcohol)) +
  geom_jitter()
```

Wine samples graded as a 9 can be counted on one hand. There's still a bit of \
overplotting. It is more clear to see the clusters of samples rise as quality \
improves.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_by_Quality_Most_Visible}
ggplot(data=wines, aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4) 
```

Reducing the alpha a bit and the change in the clusters become more clear. \
It's also clear we can't say that wine quality will be better just because the \
alcohol content is higher. This is different from saying higher quality wines \
tend to have higher alcohol content. Do not confuse this with saying lower \
quality wines will always have lower alcohol content.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_by_Quality_Categories}
ggplot(data=wines, aes(x=quality.class, y=alcohol)) +
  geom_jitter(alpha=.4)
```

Taking the bins we've defined for lesser and better quality shows a clearer \
picture. Better quality wines seems evenly dispersed across the range of \
alcohol content. It looks that the cut off is just above 10% alcohol \
content. Wines with more than 10% alcohol content are more likely to be of \
better quality.  

Considering we would rather increase than give up resolution of quality, \
it would be good take a look at the upper and lower quality wines bins from \
another perspective. To facilitate this, let's recreate the smaller bins \
without the added field.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_by_Binned_Quality}
wines %>%
  subset(as.numeric(quality) < 7) %>%
  group_by(class=cut(as.numeric(quality), breaks=seq(0, 7, by = 3))) %>%
  ggplot(data=., aes(x=class, y=alcohol)) +
  geom_jitter(alpha=.4)
```

We ended up dropping wines of quality 9. As there are so few observations with \
that rating, it could be considered outliers. Basically this is the same graph \
as previously created. The benefit of this method is allowing us to maintain \
quality granularity within wines of better of lesser quality.  

I'm also having some regret in the decision to transform quality into a \
factor. That does give more control over the scale. That additional control is \
under utilized when quality is on an integer scale.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_at_Higher_Quality}
wines %>% 
  subset(as.numeric(quality) > 3) %>%
  ggplot(aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4)
```

Taking a closer look at the breakdown of the better quality wines, We see that \
wines of quality 6 runs the full spectrum of alcohol content. Disregarding \
wines of quality 6, better wines start to occur more frequently at alcohol \
content near 10-11%. Wines of the highest quality have even fewer occurrences \
with alcohol content under ~9.5%.

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_at_Middle_Quality}
wines %>% 
  subset(as.numeric(quality) >= 3 & as.numeric(quality) <= 5) %>%
  ggplot(aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4)
```

The shift for wines of greater vs lesser quality around 10% is more apparent \
by limiting the view to the middle quality scores. Wines of quality 6 appear \
to be a merge between wines of quality 5 and 7.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_at_Lowest_Quality}
wines %>% 
  subset(as.numeric(quality) <= 3) %>%
  ggplot(aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4)
```

Breaking out the least quality wines shows a dispersion that rarely breaks \
11.5% alcohol. If the least quality wines rarely rise above 11.5% alcohol and \
the highest quality wines rarely have less than 9.5% alcohol, then the \
average quality wines must fall within that range.  

What would happen to our graphs if we eliminated the ordinary wines defined as \
having alcohol content between 9.5% and 11.5%? This also revisits the idea of \
average versus non-average wines. We have a definition of what makes for an \
ordinary wine. Eliminating those from our views, it should become easier to \
see the differences at the extremes.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_and_Quality_san-middles}
wines %>% 
  subset(alcohol <= 9.5 | alcohol >= 11.5) %>%
  ggplot(aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4) +
  stat_summary(fun.y = mean, fun.ymin = mean, fun.ymax = mean,
                 geom = "crossbar", width = 1, color = 'red')
```

The cut off between better and lesser quality wines is even easier to see when \
we remove the middle range of alcohol content. This split more clearly occurs \
at some granularity of quality rated 6.  

It's now clear that quality 6 wines are ambiguous in quality. We should look \
back at the lesser vs better quality graph while eliminating the ambiguous \
value.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Minus_Ambiguity}
wines %>%
  subset(as.numeric(quality) < 7 & as.numeric(quality) !=4) %>%
  group_by(class=cut(as.numeric(quality), breaks=seq(0, 7, by = 3))) %>%
  ggplot(data=., aes(x=class, y=alcohol)) +
  geom_jitter(alpha=.4) +
  stat_summary(fun.y = mean, fun.ymin = mean, fun.ymax = mean,
                 geom = "crossbar", width = 1, color = 'red')
```

It is now a quite clear relationship between quality and alcohol content. We \
should remember the quality scale is subjective and based on the opinion of \
experts. It's possible those experts have a preference for higher alcohol \
content in their wines. On the observations made this far, I would suggest \
that preference is likely.  

As part of due diligence, we should a look at actual value correlation between \
quality and alcohol.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Alcohol_Correlation}
cor.test(wines$alcohol, as.numeric(wines$quality))
```

Across all samples, there is a moderate correlation between quality and \
alcohol content. This correlation includes the ambiguous quality rating. This \
is a central value that fits the model but skews the data. This skew appears \
to be because of a lack of resolution in the quality scale. What happens to \
correlation if we omit this ambiguous score?  

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_to_Alcohol_Adjusted}
wines.noMiddle <- subset(wines, as.numeric(quality) != 4)
cor.test(wines.noMiddle$alcohol, as.numeric(wines.noMiddle$quality))
rm(wines.noMiddle)
```

By omitting the ambiguous quality value on our scale, the correlation \
becomes strong. I would hypothesize, if we had a more granular and less \
subjective quality scale the correlation between would be even stronger. For \
now, I will state that alcohol is an acceptable measure of quality. It lends \
greater granularity than the current quality scale.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_Summary}
summary(wines$alcohol)
```

The earlier observations on the upper limit of alcohol content for lower \
quality wines and lower limit for higher quality wines fall almost exactly on \
the 1st and 3rd quartiles of the alcohol content range. This gives even more \
confidence in the assertion that alcohol is the measure of quality.  

### Intermediate Analysis

The quality score provided by the experts lacks in granularity. Luckily, the \
experts tend to rate wines of higher alcohol content more favorably. By \
equating alcohol content with quality, we have better granularity for \
determining the composition of higher quality wines.  

On the compositional side of things, salt and sugar seem to have the most \
consistent relationship to alcohol across both varietals. There are compounds, \
such as citric acid, with stronger relationships to alcohol content. Those \
relations do not appear to be nearly as consistent across varieties.  

Moving forward in our exploration, we'll focus on alcohol, salt, sugar, and \
wine variety. The other interesting variables will be saved for future \
analysis.  

## Deeper Look

To complete our investigation, we'll start by investigating how salt \
concentration compares at various alcohol levels.  

### Salt

```{r echo=FALSE, message=FALSE, warning=FALSE, Salt_and_Alcohol}
ggplot(data=wines, aes(x=alcohol, y=chlorides)) +
  geom_jitter(alpha=.4)
```

The majority of wines have less than a hundredth of a percent of salt. There's \
a jump in salt content around 9.5 percent alcohol. Most of them appear to be \
outliers. Some scaling and subsetting should produce a better picture.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Salt_and_Alcohol_Refined}
ggplot(data=subset(wines, chlorides < .03), aes(x=alcohol, y=chlorides)) +
  geom_jitter(alpha=.4) +
  scale_y_continuous(limits = c(.002, .006)) +
  geom_smooth()
```

There's a consistent fraction of a fractional drop in salt content from \
~0.0047% to ~0.0035% as alcohol content increases from ~9% to ~13%. These \
values extend just outside the ~9.5% to ~11.5% range we previously determined \
as average quality. As wine quality increases, salt content decreases.  

What about sugar content?  

### Sugar

```{r echo=FALSE, message=FALSE, warning=FALSE, Sugar_and_Alcohol}
ggplot(data=wines, aes(x=alcohol, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_jitter(alpha=.5) +
  geom_smooth()
```

Again, the data is condensed on itself. Some adjustment on perspective can \
only help.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Sugar_and_Alcohol_Refined}
ggplot(data=subset(wines, residual.sugar <= 2.5), 
       aes(x=alcohol, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_jitter(alpha=.4) +
  geom_smooth()
```

There's a large amount of residual sugars in wines of low alcohol content. \
This seems logical. Sugar is what yeast eats to make alcohol. Less sugar \
consumed means more left over sugar and less alcohol.  

There appears to be an interesting bump in sugar content near the delineation \
of average and better than average wines. Of all the possibilities that could \
explain that bump, the one we have is variety.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Sugar_and_Alcohol_by_Varietal}
ggplot(data=subset(wines, residual.sugar <= 2.5), 
       aes(x=alcohol, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_jitter(alpha=.4) +
  geom_smooth(aes(color=color))
```

That doesn't answer anything about the bump. If anything, the bump is more \
pronounced. The interesting thing it did illustrate are the differences in \
red and white wines. It seems there's two graphs of sugar content stacked on \
each other.  

The lower graph is comprised of the red wines. Which shows a consistent level \
of residual sugar content at all levels of alcohol. In the upper graph of \
white wines, sugar drops sharply as alcohol increases before leveling out at \
the highest levels.  

### Salt Revisited

Is there a similar phenomenon in salt content?  

```{r echo=FALSE, message=FALSE, warning=FALSE, Salt_and_Alcohol_by_Varietal}
ggplot(data=subset(wines, chlorides < .03), aes(x=alcohol, y=chlorides)) +
  geom_jitter(alpha=.4) +
  scale_y_continuous(limits = c(.002, .02)) +
  geom_smooth(aes(color=color))
```

White wines have less salt than reds. White wines also have a steeper drop in \
salt content as alcohol content increases compared to red. A similar \
phenomenon except red wines tend to have more salt than white wines.  

### Sugar in Reds

Returning to sugar content, It might be helpful to break reds and whites to \
their individual scales.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Sugar_and_Alcohol_in_Reds}
ggplot(data=subset(wines, residual.sugar <= 0.7 & color == "red"), 
       aes(x=alcohol, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_jitter(alpha=.5) +
  geom_smooth(aes(color=color))
```

Unlike white wines, red wines appear to have a slight increase in sugar \
content as alcohol increases. It's so slight, starting around 0.225 and \
ending near 0.25. Given the low sample size and lack of measurement error, \
the best we can say is red wine has consistent sugar content across all \
values. There may be some miniscule differences, but there's not enough \
information to say sugar content is a contributing factor in red wine quality.

To be sure, we should check against the original quality scale.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Sugar_Quality_Reds}
ggplot(data=subset(wines, residual.sugar <= 0.7 & color == "red"), 
       aes(x=quality, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_smooth(aes(color=color)) +
  geom_boxplot()
```

The majority of values reside under 0.4% sugar content. Adjusting both graphs \
might yield a better image. Before making this adjustment, it's worth noting \
the 1st quartile and mean is fairly consistent across quality range. There is \
some variation in the 3rd quartile, but no consistent pattern.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Red_Sugar_Quality_vs_Alcohol}
p1 <- ggplot(data=subset(wines, residual.sugar <= 0.4 & color == "red"),
             aes(x=alcohol, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_jitter(alpha=.5) +
  geom_smooth(aes(color=color))

p2 <- ggplot(data=subset(wines, residual.sugar <= 0.4 & color == "red"), 
       aes(x=quality, y=residual.sugar)) + 
  scale_y_continuous() +
  geom_smooth(aes(color=color)) +
  geom_boxplot()


grid.arrange(p1,p2, ncol = 1)
```

Removing these outliers produced a more consistent pattern in both graphs. The \
sugar content for the range of alcohol content is more flat. \
Sugar content at each quality score has a consistent mean with variation \
occurring in the 1st and 3rd quartiles. 

As an aside, the similarities in these graphs reinforces our earlier assertion \
that alcohol content is a good predictor of wine quality. Unfortunately, these \
findings are contrary to our earlier observations that sugar content related \
to alcohol has a consistent relationship across varietals.  

### Putting it All Together 

One final graph I want to develop is alcohol content at ratios of salts and \
sugar.

```{r echo=FALSE, message=FALSE, warning=FALSE, Data_Selection}
wine.tastes <- data.frame("sugar" = round(wines$residual.sugar, 2), 
                          "salt" = round(wines$chlorides, 4),
                          "alcohol" = wines$alcohol,
                          "type" = wines$color)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Data_Melt}
wine.ratios <- melt(wine.tastes, id.vars = c("sugar", "salt", "type"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Beginning_of_the_End}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100)) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))
```

What we see is a concentration of higher alcohol content at lower salt and \
sugar content. Some of that could be attributed to the intersection of red and \
white wines.  

Red wine tends to have a much lower sugar content and higher salt content than \
white wine. That could explain the lower leg of this graph.  

Another thing to consider is the range of granularity alcohol content \
provides. This spread of alcohol content across the ratios of sugar to salt \
could become more apparent by introducing more colors to our graph.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Three_Colors}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradient2(low = "green", mid = "blue", high = "red", 
                       midpoint = 12) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))
```

At three colors, it seems more clear that lower alcohol content wines tend to \
occur at the extreme values of sugar and salt content.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Four_Colors}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(
    c("orange","green","blue", "red"))(50)) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))
```

At four colors, diminishing returns start to kick in. It does help solidify \
what we've already seen. There's slightly more resolution on the ratios of \
sugar and salt where the highest alcohol content wines occur. The highest \
concentration of red dots appear near 0.0025% salt and 0.25% sugar.  

It's worth noting that what has been referred to this far as a ratio between \
two compounds is a misnomer. It describes 100 grams of sugar for every gram of \
salt. We do see a higher concentration of the higher alcohol content wines \
around 0.3% sugar and 0.003% salt. We do not see any examples at 2% sugar and \
0.02% salt. There is no diagonal line that would describe a perfect ratio.  

To be more accurate, we should say that the highest alcohol content wines \
occur between 0.003% and 0.004% salt content and 0.1 to 0.75% sugar content. \
Looking closer at this range could provide additional value.  

### The Happy Spot

```{r echo=FALSE, message=FALSE, warning=FALSE, The_Happy_Spot}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradientn(colours = 
                         colorRampPalette(c("green","blue", "red"))(50)) +
  scale_y_continuous(limits = c(0,0.75)) + 
  scale_x_continuous(limits = c(0,0.006))
```

Zooming in on this range we can see that it isn't a true ratio. It's more of a \
happy spot where the balance between salt and sugar isn't too much in one \
direction or the other. There's a bit of noise at this scale with 3 colors.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Happy_Blue}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradient2(low = "white", mid = "blue", high = "black", 
                   midpoint = 11.5) +
  scale_y_continuous(limits = c(0,0.75)) + 
  scale_x_continuous(limits = c(0,0.006))
```

Gradation of one color from black to white provides a good contrast. We can \
better see where there's not enough sugar for the amount of salt in that \
bright blue. Just above the bright blue, we see a distribution of higher \
alcohol content wines extending into higher sugar content.  

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = wine.ratios) +
  geom_tile() +
  scale_fill_gradient2(low = "white", mid = "blue", high = "black", 
                   midpoint = 11.5) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))
```

Zooming back out with this color scale, the happy spot is more pronounced. \
The trade off is a loss of pronouncement in the extreme values.  

### Red vs. Whites

One thing that we've touched on previously are the composition of varietals \
and how they differ even where there's similarities.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Reds_Salt_and_Sugar}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = subset(wine.ratios, wine.ratios$type == "red")) +
  geom_tile() +
  scale_fill_gradient2(low = "white", mid = "blue", high = "black", 
                   midpoint = 11.5) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))+
  geom_smooth(aes(color=type))
```

Red wine's contribution to the happy spot is clear and not unexpected. Red \
wines have a consistent sugar content and typically have move salt. As such \
they make up the lower leg of the salt to sugar distribution.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Whites_Salt_and_Sugar}
ggplot(aes(y = sugar, x = salt, fill = value),
  data = subset(wine.ratios, wine.ratios$type == "white")) +
  geom_tile() +
  scale_fill_gradient2(low = "white", mid = "blue", high = "black", 
                   midpoint = 11.5) +
  scale_y_continuous(limits = c(0,2)) + 
  scale_x_continuous(limits = c(0,0.02))+
  geom_smooth(aes(color=type))
```

White wine's contribution to the happy spot is also quite clear. Although, \
white wine seems to be more forgiving on sugar content. Instead of saying \
white wines are more accepting of high sugar content; we should say that \
higher levels of salt need lower sugar content to remain at high quality.  
  
### Multivariate Analysis

Based on what we've seen, salt content has more impact on wine quality than \
sugar. Even with sugar rich white wines, salt content is the largest factor \
for determining quality. In white wines, you can have more sugar in a higher \
quality wine as long as the salt content remains in a minimal range.  

Between red and white wines, there is a convergence where wines of the highest \
alcohol content occurs with a balance of sugar and salt. This sweet spot \
provides some answers to the question of how wine composition relates to \
quality. What is not clear is how alcohol content and the balance of salt and \
sugar relate to quality. Are the ratios of salt and sugar the ultimate \
determinant to quality or does the alcohol content also contribute.   

Alcohol content as a granular measure of quality is something that shouldn't \
escape scrutiny. In this analysis, alcohol content was taken as a surrogate \
for quality. It offered a level of granularity not present in the quality \
scale. One issue with this is it may hide the relationship of other solutes in \
the solution and how the combination of those solutes relate to quality.  

## Final Plots and Summary

```{r echo=FALSE, message=FALSE, warning=FALSE, Building_the_Final_Plot_Theme}
thm <- theme(plot.title = element_text(size = 16, face="bold"), 
            axis.title.x = element_text(size = 10, face="bold"),
            axis.title.y = element_text(size = 10, face="bold"),
            panel.background = element_rect(fill = "seashell3", 
                                            color = "darkgrey"),
            panel.grid.major = element_line(size = .1, 
                                            color = "seashell2", 
                                            linetype = "dashed"),
            panel.grid.minor = element_line(size = .1, 
                                            color = "grey", 
                                            linetype = "dashed"),
            plot.background = element_rect(fill = "azure1"))
```

### Quality

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_As_Quality}
wines %>% 
  subset(alcohol <= 14.5) %>%
  ggplot(aes(x=quality, y=alcohol)) +
  geom_jitter(alpha=.4, shape = 21, fill = I('#2679af')) +
  stat_summary(fun.y = mean, fun.ymin = function(z){quantile(z, 0.25)}, 
               fun.ymax = function(z){quantile(z, 0.75)},
               geom = "crossbar", width = 1, fill = '#f70800', alpha = .3) +
  scale_y_continuous(breaks = seq(8, 14, 1)) +
  labs(title="Alcohol Content as Quality", 
       y="Percent Alcohol Content",
       x = "Average Quality Score") + 
  thm
```

#### Using Alcohol as Quality 

The data set has an issue of granularity within the quality scale. Without a \
a greater level of granularity, the delineation between good an bad quality \
wines is lost. Alcohol content was identified as a surrogate to quality. It \
offers a greater level of granularity.  

This granularity and logic of alcohol content as quality is illustrated here. \
The exceptionally good and exceptionally bad wines have a lack of observation \
density to produce consistency with this assumption. However, for the \
observations we do have, we see the mean of alcohol content at lesser quality \
wines are much lower than higher quality wines.  

### Salt Content of Varietals

```{r echo=FALSE, message=FALSE, warning=FALSE, Salt_in_Wine}
ggplot(data=subset(wines, chlorides < .025), 
       aes(x=alcohol, y=(chlorides*10000))) +
  geom_jitter(alpha=.3, shape = 21, fill = I('#605f5d')) +
  scale_y_continuous(limits=c(20, 120),breaks = seq(20, 120, 20)) +
  geom_smooth(data = subset(wines, color=="red"), aes(color = "Red")) +
  geom_smooth(data = subset(wines, color=="white"), aes(color = "White")) + 
  scale_colour_manual(name="Varietal", values=c("#540000", "#d2e8a9")) +
  labs(title="Salt and Alcohol Content", 
       y="Salt (ppm)", 
       x = "Alcohol (ABV)") +
  thm
```

#### Quality of Salt in Wine

Using alcohol as a granular measurement of quality revealed salt \
content as the major factor in determining wine quality. Across both \
varietals, as alcohol content increases, salt content decreases. The better \
wines have less salt.  

The amount of salt in the wine samples is miniscule. Where alcohol can be \
measured in full percents by volume, salt is better described in parts per \
million. It is interesting that slight variation in a small amount can have \
such a great effect on quality.  

We also start to see a divergence between the two main varietals of wine. Red \
wines tend to have almost twice the amount of salt of white wines. This \
suggests that quality is reliant on a combination of compounds.  


### Alcohol, Salt and Sugar Content

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(y = sugar*1.48, x = salt*1480),
  data = wine.ratios) +
  geom_raster(aes(fill=value),interpolate = TRUE) +
  scale_fill_gradientn(colors = brewer.pal(n = 5,name = "Greys")) +
  scale_y_continuous(limits = c(0,3.3)) + 
  scale_x_continuous(limits = c(0,30)) +  
  geom_smooth(data = subset(wine.ratios, type=="red"), aes(color = "Red")) +
  geom_smooth(data = subset(wine.ratios, type=="white"), aes(color = "White")) + 
  scale_colour_manual(name="Varietal", values=c("#540000", "#d2e8a9")) +
  labs(title="ABV Related to Salt and Sugar", y="Sugar per Serving (g)",
       x = "Salt per Serving (mg)", fill = "ABV") +
  thm
```

#### Composition of Varietal Quality

We now start to see the relationship between combinations of compounds and \
quality. Also apparent are the differences and similarities between red and \
white wines. There's a convergence between both varietals for ratios of salt \
and sugar where the highest alcohol content occurs. This sweet spot for salt \
and sugar is represented in the darkened area of the graph.  

There's a range of salt that is acceptable for reds and whites. There's a \
similar range for sugar as well. The difference for varietals is which solute \
is more forgiving in quality. More salt is acceptable in red wines as long as \
it remains within a limitation of sugar. The inverse is true for white wines.  

## Reflection

One thing that strikes me about this data is how worthless the quality score \
is. Asking someone to rate something on a scale of one to whatever does offer \
some insight. The question here is, how many opinions using that scale is \
needed to get a usable granularity? It's a great starting point with limited \
use.  

Despite the limitations of quality score, the dataset does offer a depth of \
possibilities. Differences of composition between varietals absent of \
quality considerations is the top of my list. While alcohol content was \
chosen as a surrogate to quality, it still is a measure of alcohol content. It \
could be that alcohol is a good measure of quality. I would argue that the \
polled experts have a bias for alcohol content. What makes them experts and \
what criteria do they use in rating decisions?  

There's also many items noted in the initial pass of the data that were not \
explored further. One compound not explored and deserving a future analysis \
are the acids. There's enough evidence to explore the relationship between \
citric acid, sugar and salt. There's also an expectation that Ph is influenced \
by acidic compounds. Expectations should always be tested.  

At the end of it, the gleaned insight is something I look forward to testing. \
Next time I'm in the store for a bottle of wine, I'm going to use alcohol \
content as my primary deciding factor. It will be interesting how many bottles \
go through before it proves to be a bad metric.  